---
title: "plyxp: From Tidy Code to Tidy Transcripts"
---

```{r}
source("R/zzz-load.R")

.render_clone <- NULL

render_ <- function(obj, file) {
  .render_clone <<- obj_clone(obj)
  render_gif(obj, file, width = 1400, height = 700, loop = 1,
    fps = 25, clone = FALSE, 
    units = "px")
}


```

```{r}


w_w <- 10
w_h <- 5
w <- window(bl = 0, tr = pos(w_w, w_h))
trk <- tracker(w)
.c <- center(w)
.size <- w_h * 0.9

.assay <- rect(width = .size, height = .size, color = "#8BD79188",
  trans = transform(.c))


w@child(
  .assay@action(
    spawn_grid(
      nrow = 10, ncol = 8, color = "#8BB85F",
      padding = 0.15,
      xtime = 0.1,
      apply = \(x) {obj_scale(x, size = -1)},
      action = list(scale2(size = 1.2, time = time(.7))@
       then(scale2(size = -.2, time = time(.2))))
    )
  )
)

trk@track(.assay = .assay)

render_(w, "assay_example.gif")
bk0 <- trk@clone()
```

```{r}
trk <- bk0@restore()
w <- trk@object
pal <- scales::pal_gradient_n(scales::pal_viridis(1, 0, 1, 1, "D")(6))

set.seed(123)
lapply(.assay@children, \(child) {
  child$start <- child@color
  child$val <- runif(1)
  child$end <- pal(child$val)
  child@action(
    color(child$start, child$end, time = time(3, ease = ease(ecubic)))
  )
}) |> invisible()


render_(w, "assay_color.gif")
bk1 <- trk@clone()
```

```{r}


scale_top <- function(size, time) {
  time <- new_time(time)
  action(function(obj, time) {
    force(time)
    if (is.null(time$from)) {
      time$from <- obj_size(obj)
      .clone <- obj_scale(obj_clone(obj), size = size)
      time$width <- .clone@width
      time$height <- .clone@height
      time$around <- tc(.clone)
    }
    from <- time$from
    height <-  time$height * time@value
    obj@height <- height
    obj@width <- time$width
    obj_translate(obj, pos(time$around$x, time$around$y - (height / 2)))
  }, time = time)
}
```

```{r}
trk <- bk1@restore()
w <- trk@object

.lc <- lc(.assay)
.rows <- rect(height = .assay@height, width = .lc$x * 0.8,
  transform(
    pos(mean(c(0, .lc$x)), .lc$y) 
  ),
  color = "#AC94CA00"
) 
w@child(.rows)
trk@track(.rows = .rows)

.rows@action(
  color(
    "#AC94CA00", "#AC94CAFF", time = time(1, ease = ecubic)
  )@then(
    spawn_grid(nrow = 1, ncol = 3, xtime = 0, 
      color = "#AC43CE", ypadding = .02, xpadding = .1,
      apply = \(x) obj_scale(x, size = -1),
    actions = list(
      scale_top(size = 1, 
        time = time(.4, ease = ease(ecubic, ecubic))
      )@then(
        spawn_grid(nrow = 10, ncol = 1, xtime = 0,
          color = "#AC6ACA", padding = .2,
          apply = \(x) obj_scale(x, size = -1),
          action = list(scale2(size = 1, time = time(.7)))
        )
      )))
  )
)

render_(w, "assay_rows.gif" )
bk2 <- obj_clone(trk)
```

```{r}
trk <- bk2@restore()
w <- trk@object

```

```{r}
trk <- bk2@restore()
w <- trk@object
.bl <- bl(.rows)

w@action(
  action(
    function(obj, time) {
      obj_scale(obj@children[[1]], .bl, size = -0.25 * time@delta)      
      obj_scale(obj@children[[2]], .bl, size = -0.25 * time@delta)
    },
    time = time(1.5, ease = ease(ecubic, ecubic))
  )
)
render_(w, "shrink_assay_row.gif")
bk3 <- trk@clone()
```


```{r}
trk <- bk3@restore()
w <- trk@object
"#FFDC8F" #inner
"#FFA909" #column
"#E1C757" #context
.tc <- tc(.assay)
.cols <- rect(height = (5 - .tc$y) * .8 , width = .assay@width,
  transform(
    pos(.tc$x, mean(c(5, .tc$y))) 
  ),
  color = "#E1C75700"
) 
w@child(.cols)
trk@track(.cols = .cols)

.cols@action(
  color(
    "#E1C75700", "#E1C757FF", time = time(1, ease = ecubic)
  )@then(
    spawn_grid(nrow = 2, ncol = 1, xtime = 0, 
      color = "#FFA909", xpadding = .015, ypadding = .1,
      apply = \(x) obj_scale(x, size = -1),
    actions = list(
      scale_top(size = 1, 
        time = time(.4, ease = ease(ecubic, ecubic))
      )@then(
        spawn_grid(nrow = 1, ncol = 8, ytime = 0,
          color = "#FFDC8F", padding = .2, xpadding = .15,
          apply = \(x) obj_scale(x, size = -1),
          action = list(scale2(size = 1, time = time(.7)))
        )
      )))
  )
)

render_(w, "assay_cols.gif" )
bk4 <- trk@clone()
```

```{r}
trk <- bk4@restore()
w <- trk@object

# reorder children elements position in world space
.assay@children <- .assay@children[lapply(.assay@children, obj_pos) |> vctrs::list_unchop() |> (\(x) order(x$x, x$y, decreasing = c(FALSE, TRUE)))()]

.df <- rect(width = 4, height = 5, 
  trans = transform(pos(7.8, 2.5)),
  color = "grey80")

.dest_cache <- calc_grid(
  .df, nrow = 80, ncol = 6, 
  xpadding = 0.01,
  xmargin = 1
)
.height <- attr(.dest_cache, "features")["height"]
.width <- attr(.dest_cache, "features")["width"]

.simple_scale <- function(obj, time) {
  if (is.null(time$dh)) {
    time$dh <- .height - obj@height
    time$dw <- .width - obj@width 
  }
  obj@height <- obj@height + (time$dh * time@delta)
  obj@width <- obj@width + (time$dw * time@delta)
}
w@child(.df)
trk@track(.df = .df)
```

```{r}
.time_sc <- NULL
.assay@action(
  action(
    function(obj, time) {
      if (time@cycled) {
        if (is.null(time$index)) {
          time$index <- 0L
        }
        time$index <- time$index + 1L
        .time_sc <<- time@time_scale <- time@time_scale * 1.01 # accelerate
        orig_child <- obj@children[[time$index]]
        child <- obj_clone(orig_child)@action(
          translate(.dest_cache[[1]][time$index], time = time(.75, ease = ease(ecubic))),
        )@action(
          .simple_scale,
          time = time(.75, ease = ease(ecubic))
        )
        orig_child@color <- opaque(orig_child@color, 0.5)
        spawn_from(.df, child)
      }
    },
    time = time(.1, repeating = 79)
  )
)

render_(w, "unpacking_assay.gif")
bk5 <- trk@clone()
```

```{r}
trk <- bk5@restore()
w <- trk@object
.rtime_sc <- NULL
.unpack_row <- function(obj, time) {
  if (time@cycled) {
    if (is.null(time$col)) {
      time$col <- 1L
      time$i <- 0L
    }
    if (is.null(time$index) || time$index == 10L) {
      time$index <- 0L
    }
    if (time$i == 80) {
      time$i <- 0L
      time$col <- time$col + 1L
    }
    time$i <- time$i + 1L
    time$index <- time$index + 1L
    .rtime_sc <<- time@time_scale <- min(time@time_scale * 1.01, 70) # accelerate
    column <- obj@children[[time$col]]
    orig_child <- column@children[[time$index]]
    child <- obj_clone(orig_child)@action(
      translate(.dest_cache[[time$col + 1L]][time$i],
        time = time(.75, ease = ease(ecubic))),
    )@action(
      .simple_scale,
      time = time(.75, ease = ease(ecubic))
    )
    orig_child@color <- opaque(orig_child@color, 
      opaqueness(orig_child@color) - 0.0625)
    spawn_from(.df, child)
  }
}

.rows@action(
  action(
    function(obj, time) {
      if (time@cycled) {
        time@reset()
      }
      if (length(.assay@actions) == 0L) {
        time@cycled <- TRUE
        time@iter <- time@iter + 1L
      }
    },
    # inf time is conditional
    time = time(1)
  )@then(
    action(
      .unpack_row,
      time = time(1 / .time_sc, repeating = 239)
    )
  )
  
)

render_(w, "unpacking_rows.gif")
bk6 <- trk@clone()
```

```{r}
trk <- bk6@restore()
w <- trk@object
.unpack_col <- function(obj, time) {
  if (time@cycled) {
    if (is.null(time$col)) {
      time$col <- 1L
      time$i <- 0L
    }
    if (is.null(time$index)) {
      time$index <- 0L
    }
    if (time$i == 80) {
      time$i <- 0L
      time$col <- time$col + 1L
      time$index <- 0L
    }
    
    
    time$i <- time$i + 1L
    
    if ((time$i - 1L) %% 10L == 0L) {
      time$index <- time$index + 1L
    }
    
    time@time_scale <- min(time@time_scale * 1.01, 70) # accelerate
    column <- obj@children[[time$col]]
    orig_child <- column@children[[time$index]]
    child <- obj_clone(orig_child)@action(
      translate(.dest_cache[[time$col + 4L]][time$i],
        time = time(.75, ease = ease(ecubic))),
    )@action(
      .simple_scale,
      time = time(.75, ease = ease(ecubic))
    )
    orig_child@color <- opaque(orig_child@color, 
      opaqueness(orig_child@color) - 0.0625)
    spawn_from(.df, child)
  }
}

.cols@action(
  action(
    function(obj, time) {
      if (time@cycled) {
        time@reset()
      }
      if (length(.rows@actions) == 0L) {
        time@cycled <- TRUE
        time@iter <- time@iter + 1L
      }
    },
    # inf time is conditional
    time = time(1)
  )@then(
    action(
      .unpack_col,
      time = time(1/40, repeating = 159)
    )
  )
  
)

render_(w, "unpacking_cols.gif")
bk7 <- trk@clone()
```

```{r}
local({
  .g1 <- magick::image_read("unpacking_assay.gif")
  .g2 <- magick::image_read("unpacking_rows.gif")
  .g3 <- magick::image_read("unpacking_cols.gif")
  magick::image_animate(c(.g1, .g2, .g3), fps = 25, loop = 1) |>
  magick::image_write("unpacking.gif")
})

```

```{r}
.assay <- .w@children[[1]]
.assay@children <- .assay@children[lapply(.assay@children, obj_pos) |>
vctrs::list_unchop() |> order()]
.pos <- vctrs::list_unchop(lapply(.assay@children, obj_pos))
row_order <- vapply(.assay@children, \(x) x$val, 1) |>
  matrix(10, 8) |>
  dist() |>
  hclust() |>
  _$order
xtent <- get_extent(.assay@children)
target_y <- matrix(vctrs::field(.pos, "y"), 10, 8) |> _[row_order,]
col_order <- vapply(.assay@children, \(x) x$val, 1) |>
  matrix(10, 8) |>
  t() |>
  dist() |>
  hclust() |>
  _$order
target_x <- matrix(vctrs::field(.pos, "x"), 10, 8) |> _[,col_order]
mapply(
  \(child, target) {
    child$old_pos <- child@trans@global
    child$new_pos <- target
    child@action(
      translate(target, time = time(3))
    )
  },
  child = .assay@children,
  target = pos(as.vector(target_x), as.vector(target_y))
)
.w3 <- anim(.w)

# mapply(
#   \(child, target) {
#     child$new_y <- target
#     child@action(
#       offset(by = pos(0, target - vctrs::field(obj_pos(child), "y")), time = time(3))
#     )
#   },
#   child = .w@children,
#   target = vctrs::vec_rep(target_y, 8)
# )
# # .w2 <- anim(.w)
# mapply(
#   \(child, target) {
#     child$new_y <- target
#     child@action(
#       offset(by = pos(target - vctrs::field(obj_pos(child), "x"), 0), time = time(3))
#     )
#   },
#   child = .w@children,
#   target = vctrs::vec_rep_each(target_x, 10)
# )

```

## Introduction

my name is Justin Landis and today I get to share some of the work I have done since joining the Love Lab.

### Overview

To somewhat spoil the contents of this presentation, I will primarily be discussing the R package I have developed pronounced 'plyxp'.
In a way, this student seminar works well for me, as I have a personal goal of advertising this package to its target audience.
And while not everyone in this room may use Bioconductor, or even R packages, this particular room has a higher chance to be an interested user than say any random individual in chapel hill.
That being said, the work I will be presenting on today is currently being drafted into an application note, where we have made an intentional effort to describe the framework as language agnositic as possible.
So if by the end of the talk you would be interested in maybe expanding this to python, please reach out to me and we can grab a coffee.

### What is tidyomics



## Timeline

May 2024 - tidyomics introduced
Oct 2024 - plyxp accepted into Bioconductor
June 2025 - GBCC talk
August 2025 - UseR!: tidyomics workshop
Oct 2025 - tidySumamrizedExperiment started to use plyxp
Oct 2025 - CZI project showcase
2026? - plyxp backend released as biocmask
2026? - use biocmask to rewrite plyranges?


## plyxp syntax

## plyxp semantics

## call for interest in python

## fast forward to transcripts (eventual goal)

## show case progress
