---
title: "Getting Started with ranim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ranim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 8,
  fig.height = 6
)
library(ranim)
```

## Introduction

**ranim** is an R package for creating hierarchical, shape-based animations. Shapes can contain other shapes and can be manipulated over time with composable "action" objects that drive animation. The package is designed around the concept of a tree hierarchy: each shape can have a parent and children, and transformations applied to a parent automatically affect all descendants.

### Key Concepts

1. **Shapes**: The fundamental building blocks. Includes `rect()`, `point()`, `polygon()`, `text()`, and `line()`.
2. **Hierarchy**: Shapes can be nested; a child's position and size are relative to its parent.
3. **Actions**: Time-varying functions that modify shapes (e.g., `translate()`, `rotate()`, `scale2()`, `color()`).
4. **Time**: Controls animation pacing, easing, and repetition via `time()` objects.
5. **Windows**: Define the canvas for animation via `window()`.
6. **Rendering**: Display animations on-screen with `anim()` or export to GIF with `render_gif()`.

## Creating a Simple Animation

### Step 1: Create a Window

```{r}
library(ranim)

# Create a 10x5 window (canvas)
w <- window(bl = pos(0, 0), tr = pos(10, 10))
```

### Step 2: Add Shapes

```{r}
# Create a rectangle at the center
rect_obj <- rect(
  width = 2, 
  height = 2, 
  color = "red",
  trans = transform(pos(5, 5))
)

# Add it as a child of the window
w@child(rect_obj)
```

### Step 3: Add Actions

```{r}
# Make the rectangle move to a new position over 2 seconds
rect_obj@action(
  translate(to = pos(8, 8), time = time(2))
)
```

### Step 4: Animate

```{r}
# Animate on screen at 25 FPS
anim(w, fps = 25)
```

Or export to a GIF file:

```{r}
# Export to GIF
render_gif(w, "animation.gif", fps = 25, width = 8, height = 6, loop = 0)
```

## Working with Hierarchies

One of ranim's strengths is managing shape hierarchies. Children inherit transformations from their parents:

```{r}
w <- window(bl = pos(0), tr = pos(10))

# Create a parent group
group <- shape(trans = transform(pos(5, 5)))
w@child(group)

# Add several children to the group
for (i in 1:3) {
  child <- point(
    color = c("red", "blue", "green")[i]
  )
  group@child(child, offset = pos(i - 2, 0))
}

# Animate the entire group
group@action(
  rotate(degrees = 360, time = time(3))
)

anim(w)
```

## Action Composition

Actions can be chained together using `@then()`:

```{r}
w <- window(bl = pos(0), tr = pos(10, 10))
rect_obj <- rect(width = 2, height = 2, trans = transform(pos(5, 5)))
w@child(rect_obj)

# Chain multiple actions
rect_obj@action(
  translate(to = pos(8, 8), time = time(0.5))@then(
    translate(to = pos(2, 2), time = time(1))
  )@then(
    color("red", "blue", time = time(1))
  )
)

anim(w)
```

## Easing Functions

Control how animations progress over time with easing functions:

```{r}
# Linear easing (default)
t1 <- time(duration = 1, ease = elinear)

# Cubic ease-in
t2 <- time(duration = 1, ease = ease(ein = ecubic))

# Cubic ease-in-out
t3 <- time(duration = 1, ease = ease(ein = ecubic, eout = ecubic))

# Custom: blend cubic in and sine out
t4 <- time(duration = 1, ease = ease(ein = ecubic, eout = esin))
```

The `ease()` class comes with a plotting generic to visualize what the curve will be over time.

```{r}
plot(t1@ease)
plot(t2@ease)
plot(t3@ease)
plot(t4@ease)
```

Use them in actions:

```{r}
w <- window(bl = pos(0), tr = pos(10, 10))
rect_obj <- rect(width = 2, height = 2, trans = transform(pos(5, 5)))
w@child(rect_obj)

rect_obj@action(
  translate(to = pos(8, 8), time = time(1, ease = ease(ein = ecubic)))
)

anim(w)
```

## Spawning Grids

The `spawn_grid()` action creates a grid of child shapes with optional staggering:

```{r}
w <- window(bl = pos(0), tr = pos(10, 10))
container <- rect(
  trans = transform(pos(5, 5)),
  color = opaque("white", 0),
  height = 4, width = 4
)
w@child(container)

# Create a 5x5 grid of rectangles
container@action(
  spawn_grid(
    nrow = 5, ncol = 5,
    color = "steelblue",
    xtime = 0.1, ytime = 0.1,  # Stagger timing
    actions = list(
      scale2(size = -0.5, time = time(0.5))
    )
  )
)

anim(w, fps = 20)
```

## Using Trackers

Since most objects are environments, this can make iterating difficult to do. The `tracker()` object helps manage complex hierarchies by preserving references to specific shapes.



```{r}
w <- window(bl = pos(0), tr = pos(10, 10))
trk <- tracker(w)

rect1 <- rect(width = 2, height = 2, trans = transform(pos(3, 5)))
rect2 <- rect(width = 2, height = 2, trans = transform(pos(7, 5)))
w@child(rect1)
w@child(rect2)

# Track specific shapes
trk@track(rect1 = rect1, rect2 = rect2)

# Add animations
rect1@action(color("red", "blue", time = time(2)))
rect2@action(color("green", "yellow", time = time(2)))

# Save a checkpoint
checkpoint <- trk@clone()

# Continue animating
rect1@action(translate(to = pos(5, 5), time = time(1)))
rect2@action(translate(to = pos(5, 5), time = time(1)))

anim(w)

# Later, restore to checkpoint
restored <- checkpoint@restore()

# Continue animating
rect1@action(translate(to = pos(7, 5), time = time(1)))
rect2@action(translate(to = pos(3, 5), time = time(1)))

anim(restored@object)
```

## Advanced: Custom Actions

Create custom actions using the `action()` constructor:

```{r}
# Define a custom spiral motion
spiral_action <- function(obj, time_duration) {
  t <- time(duration = time_duration)
  action(
    function(obj, time) {
      # Move in a spiral pattern
      angle <- 2 * pi * time@value
      radius <- 3 * time@value
      x <- 5 + radius * cos(angle)
      y <- 5 + radius * sin(angle)
      obj_translate(obj, to = pos(x, y))
    },
    time = t
  )
}

w <- window(bl = pos(0), tr = pos(10, 10))
rect_obj <- rect(width = 1, height = 1, trans = transform(pos(5, 5)))
w@child(rect_obj)

rect_obj@action(spiral_action(rect_obj, 3))
anim(w)
```

## Color Animations

Smoothly transition between colors:

```{r}
w <- window(bl = pos(0), tr = pos(10))
rect_obj <- rect(width = 2, height = 2, color = "red", trans = transform(pos(5, 5)))
w@child(rect_obj)

# Single color transition
rect_obj@action(
  color("red", "blue", time = time(2))
)

anim(w)
```

Multi-color sequence:

```{r}
w <- window(bl = pos(0), tr = pos(10))
rect_obj <- rect(width = 2, height = 2, trans = transform(pos(5, 5)))
w@child(rect_obj)

# Cycle through multiple colors
rect_obj@action(
  colors(c("red", "purple", "blue", "green", "yellow", "red"), durations = 1, repeating = 2)
)

anim(w)
```

## Summary

The ranim package provides:

- **Hierarchical shapes** for complex scene composition
- **Composable actions** for flexible animation control
- **Flexible timing** with easing functions and repetition
- **Multiple rendering modes**: interactive (`anim()`), batch (`render_gif()`)
- **Checkpoint system** (`tracker()`) for managing animation states

For more information, see the function documentation:

- `?shape` — Base shape class
- `?window` — Canvas for animation
- `?action` — Animation actions
- `?time` — Time control
- `?translate`, `?rotate`, `?scale2`, `?color` — Common actions
- `?render_gif` — Export to GIF
